Software Citation Principles

### 2016-03-22 sc-principles/software-citation-principles.tex
some formatting cleanup in various locations

In a blog post from 2012, Jackson discusses some of the pitfalls of trying to cite software in publications~\cite{ssi-how-to-cite}.
He includes useful guidance for when to consider citing software as well as some ways to help `convince' journal editors
to allow the inclusion of software citations.

Wilson suggests that software authors include a `CITATION' file that documents exactly how the
authors of the software would like to be cited by others~\cite{ssi-citation-files}. While this isn't a formal metadata specification (it's not machine
readable for example) this does offer a solution for authors wishing to give explicit instructions to potential citing authors and as noted in the introduction (\S\ref{sec:intro}), there is evidence that authors follow instructions if they exist~\cite{10.1371/journal.pone.0136631}.

In a later post on the SSI blog, Jackson gives a good overview of some of the approaches
package authors have taken to automate the generation of citation entities such as bibtex entries~\cite{ssi-how-shalt-i-cite-thee},
and Knepley et al.\ do similarly~\cite{knepley2013accurately}.

Perhaps in recognition of the broad range of research domains struggling with
the challenge of better recognizing the role of software, a number of community
efforts hosted (and sponsored) by funders and agencies in both the US (e.g., NSF,
NIH, Alfred P. Sloan Foundation) and UK (e.g., SFTC, JISC) have run a number of
workshops with participants from across a range of disciplines.

@@

One approach to increasing the `citability' of software is to encourage the
submission of papers in standard journals describing a piece of research
software, often known as software papers (see \S\ref{sec:software_papers}).
While some journals (e.g., Transactions on Mathematical Software [TOMS], Bioinformatics, Computer Physics Communications, F1000Research) have traditionally accepted
software submissions, the American Astronomical Society (AAS) have recently
announced they will accept software papers in their journals~\cite{aas-sofware-papers}. 
Professional societies are in a good position to change their respective communities, as the publishers of journals and conveners of domain-specific conferences; as publishers they can change editorial policies (as AAS has done) and conferences are an opportunity to communicate and discuss these changes with their communities.

@@

The Unique identification principle (\ref{principle:uid}) calls for ``a method for identification that is machine actionable, globally unique, interoperable, and recognized by a community.''
What this means for data is discussed in detail in the `Unique Identification' section of a report by the FORCE11 Data Citation Implementation Group (DCIG)~\cite{10.7717/peerj-cs.1}, which calls for ``unique identification in a manner that is machine-resolvable on the Web and demonstrates a long-term commitment to persistence.''
This report also lists examples of identifiers that match these criteria including DOIs, PURLs, Handles, ARKS, and NBNs.
For software, we recommend the use of DOIs as the unique identifier due to their common usage and acceptance, particularly as they are the standard for other digital products such as publications.

>>>>

In a blog post from 2012, Jackson discusses some of the pitfalls of trying to cite software in publications~\cite{ssi-how-to-cite}.
He includes useful guidance for when to consider citing software as well as some ways to help ``convince'' journal editors
to allow the inclusion of software citations.

Wilson suggests that software authors include a \texttt{CITATION} file that documents exactly how the authors of the software would like to be cited by others~\cite{ssi-citation-files}. While this is not a formal metadata specification (e.g., it is not machine readable) this does offer a solution for authors wishing to give explicit instructions to potential citing authors and as noted in the introduction (\S\ref{sec:intro}), there is evidence that authors follow instructions if they exist~\cite{10.1371/journal.pone.0136631}.

In a later post on the SSI blog, Jackson gives a good overview of some of the approaches
package authors have taken to automate the generation of citation entities such as \textsc{Bib}\TeX\ entries~\cite{ssi-how-shalt-i-cite-thee},
and Knepley et al.\ do similarly~\cite{knepley2013accurately}.

Perhaps in recognition of the broad range of research domains struggling with
the challenge of better recognizing the role of software, a number of community
efforts hosted (and sponsored) by funders and agencies in both the US (e.g., NSF,
NIH, Alfred P.\ Sloan Foundation) and UK (e.g., SFTC, JISC) have run a number of
workshops with participants from across a range of disciplines.

@@

One approach to increasing the ``citability'' of software is to encourage the
submission of papers in standard journals describing a piece of research
software, often known as software papers (see \S\ref{sec:software_papers}).
While some journals (e.g., Transactions on Mathematical Software [TOMS], Bioinformatics, Computer Physics Communications, F1000Research) have traditionally accepted
software submissions, the American Astronomical Society (AAS) have recently
announced they will accept software papers in their journals~\cite{aas-sofware-papers}. 
Professional societies are in a good position to change their respective communities, as the publishers of journals and conveners of domain-specific conferences; as publishers they can change editorial policies (as AAS has done) and conferences are an opportunity to communicate and discuss these changes with their communities.

@@

The Unique identification principle (\ref{principle:uid}) calls for ``a method for identification that is machine actionable, globally unique, interoperable, and recognized by a community.''
What this means for data is discussed in detail in the ``Unique Identification'' section of a report by the FORCE11 Data Citation Implementation Group (DCIG)~\cite{10.7717/peerj-cs.1}, which calls for ``unique identification in a manner that is machine-resolvable on the Web and demonstrates a long-term commitment to persistence.''
This report also lists examples of identifiers that match these criteria including DOIs, PURLs, Handles, ARKS, and NBNs.
For software, we recommend the use of DOIs as the unique identifier due to their common usage and acceptance, particularly as they are the standard for other digital products such as publications.
